Modalities for Virtualizing Memory Addresses

### 2022-09-25 cpp_paper.tex
some todos and some notes about Kolanski and Klein
new format
+

%%
%% This is file `sample-sigplan.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `sigplan')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigplan.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass command.

\documentclass[sigplan,10pt,anonymous,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
\usepackage[utf8]{inputenc}
\usepackage{todonotes}
\usepackage{blindtext}
\usepackage{listings}
\lstset{
  columns=fullflexible,
  numbers=left,
  basicstyle=\ttfamily,
  keywordstyle=\color{blue}\bfseries,
  morekeywords={mov,add,call},
  emph={rsp,rdx,rax,rbx,rbp,rsi,rdi,rcx,r8,r9,r10,r11,r12,r13,r14,r15},
  emphstyle=\color{green},
  emph={[2]cr3},
  emphstyle={[2]\color{violet}},
  morecomment=[l]{;;}
}
% Hack for Colin to edit in low light: https://tex.stackexchange.com/questions/40495/invert-background-and-text-colours-across-whole-document-with-pdflatex
%Commands definitions
\newcommand{\setbackgroundcolour}{\pagecolor[rgb]{0.19,0.19,0.19}}  
\newcommand{\settextcolour}{\color[rgb]{0.77,0.77,0.77}}    
\newcommand{\invertbackgroundtext}{\setbackgroundcolour\settextcolour}
%Command execution. 
%If this line is commented, then the appearance remains as usual.
\invertbackgroundtext

@@

%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[Conference acronym 'XX]{Make sure to enter the correct
  conference title from your rights confirmation emai}{June 03--05,
  2018}{Woodstock, NY}
\acmPrice{15.00}
\acmISBN{978-1-4503-XXXX-X/18/06}

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Modalities for Location Virtualization}
\subtitle{Virtualizing Memory Addresses}
%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{X X}
\authornote{Both authors contributed equally to this research.}
\email{trovato@corporation.com}
\orcid{1234-5678-9012}
\author{G.K.M. Tobin}
\authornotemark[1]
\email{webmaster@marysville-ohio.com}
\affiliation{%
  \institution{Institute for Clarity in Documentation}
  \streetaddress{P.O. Box 1212}
  \city{Dublin}
  \state{Ohio}
  \country{USA}
  \postcode{43017-6221}
}

@@

\input{intro.tex}
%\input{motivation.tex}
\input{background.tex}
\input{syntax.tex}
\input{semantics.tex}
\input{programlogic.tex}
\input{soundness.tex}
\input{examples.tex}
\input{discussion.tex}
\input{relwork.tex}
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{vmm}

>>>>

%%
%% This is file `sample-acmsmall-conf.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall-conf')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall-conf.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[acmsmall]{acmart}

%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[Conference acronym 'XX]{Make sure to enter the correct
  conference title from your rights confirmation emai}{June 03--05,
  2018}{Woodstock, NY}
%%
%%  Uncomment \acmBooktitle if the title of the proceedings is different
%%  from ``Proceedings of ...''!
%%
%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
%%  June 03--05, 2018, Woodstock, NY}
\acmPrice{15.00}
\acmISBN{978-1-4503-XXXX-X/18/06}

\usepackage[utf8]{inputenc}
\usepackage{todonotes}
\usepackage{blindtext}
\usepackage{listings}
\lstset{
  columns=fullflexible,
  numbers=left,
  basicstyle=\ttfamily,
  keywordstyle=\color{blue}\bfseries,
  morekeywords={mov,add,call},
  emph={rsp,rdx,rax,rbx,rbp,rsi,rdi,rcx,r8,r9,r10,r11,r12,r13,r14,r15},
  emphstyle=\color{green},
  emph={[2]cr3},
  emphstyle={[2]\color{violet}},
  morecomment=[l]{;;}
}
% Hack for Colin to edit in low light: https://tex.stackexchange.com/questions/40495/invert-background-and-text-colours-across-whole-document-with-pdflatex
%Commands definitions
\newcommand{\setbackgroundcolour}{\pagecolor[rgb]{0.19,0.19,0.19}}  
\newcommand{\settextcolour}{\color[rgb]{0.77,0.77,0.77}}    
\newcommand{\invertbackgroundtext}{\setbackgroundcolour\settextcolour}
%Command execution. 
%If this line is commented, then the appearance remains as usual.
%\invertbackgroundtext

@@

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Modalities for Virtualizing Memory Addresses}
\subtitle{Understanding the Locality with Modals}
%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{X X}
\authornote{Both authors contributed equally to this research.}
\email{trovato@corporation.com}
\orcid{1234-5678-9012}
\author{G.K.M. Tobin}
\authornotemark[1]
\email{webmaster@marysville-ohio.com}
\affiliation{%
  \institution{Institute for Clarity in Documentation}
  \streetaddress{P.O. Box 1212}
  \city{Dublin}
  \state{Ohio}
  \country{USA}
  \postcode{43017-6221}
}

@@

\todo[inline]{Colin think through instruction pointer updates vs logic,
think about framing plus cr3 update and modality shifts
}

\input{intro.tex}
%\input{motivation.tex}
\input{background.tex}
\input{syntax.tex}
\input{semantics.tex}
\input{programlogic.tex}
\input{soundness.tex}
\input{examples.tex}
\input{discussion.tex}
\input{relwork.tex}
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{vmm}
\begin{appendix}
\section{Complete Syntax}
\label{appendix:syntax}
\begin{figure}[!ht]
\newcommand{\commentary}[1]{ & \text{\small\it #1} \\}
\[
  \begin{array}{r@{\;}c@{\;}l}
    \loc & \in & \Loc \\
    \reg & \in & \regset \\
    \regval & \in & \regvaltype \\
    \val & ::= & \vunit

\\
    \instrs & ::= &
    \begin{array}[t]{@{}l@{\hspace{10mm}}l@{}}
    \begin{array}[t]{@{}ll@{}}
      \iskip
                   \commentary{no-op}
      \iseq\instr\instrs
                   \commentary{sequencing}
      \ising\instr
                   \commentary{executing}             
    \end{array}
    %&
    %\begin{array}[t]{@{}ll@{}}
     % \ialloc\lval\allocsize
      %             \commentary{heap allocation}
    %\end{array}
    \end{array}
    \\

    \ectx & ::= &
      \hole \mid
      \iseq\ectx\instrs 
    \\
  \end{array}
\]
\caption{Syntax}
\Description{Syntax}
\label{fig:syntax}
\end{figure}
\section{Operational Semantics of Selected Mov Instructions}
\label{appendix:movops}
\input{fig-semantics}

\end{appendix}

### 2022-09-25 programlogic.tex
+

\section{Program Logic for Location Virtualization}
\label{sec:logic}
In this section, we describe our high-level reasoning principles for location virtualization whose detailed justification is shown in \sref{sec:soundness}. Here we present our reasoning principle shaped around understanding the virtual address space, specifially the opeations using virtual points-to mappings.
\subsection{Points-To Assertions}
\label{sec:pointsto}

>>>>

\section{Program Logic for Location Virtualization}
\label{sec:logic}
In this section, we describe our high-level reasoning principles for location virtualization whose detailed justification is shown in \sref{sec:soundness}. The principles shown here are mainly shaped around understanding and expressing the fact related the virtual address space, most specifically virtualization and translation of memory locations .
\subsection{Points-To Assertions}
\label{sec:pointsto}

### 2022-09-25 semantics.tex
+

\section{Semantics}
\label{sec:semantics}
\subsection{State}
\label{sec:state}
We represent the machine state mainly as a finite map of registers to register values and a map of masked physical memory addresses to 64-bit physical memory values. Together with default def\textsf{CPU} instantiation, we end up having our state with following pieces:
\begin{itemize}
\item A constant cpu initialized with default value: $\sigma.\mathcal{C}$
\item A register map: $\sigma.\mathcal{R}: \kw{greg} \rightarrow_{\textrm{fin}} \kw{regval} $
\item A memory map: $\sigma.\mathcal{M}: \Locft \rightharpoonup_{\textrm{fin}} (\Loctw \rightharpoonup_{\textrm{fin}} \Locsf )$
\end{itemize}
\subsection{Indexed Memory \& Register by References}
\label{sec:}
As one might have already anticipated from the syntax we introduce, we do not bind any value of an evaluated expression. All the indices and accessed values are treated as globally referenced. In align with this design choice, our expression is a stream of instructions, which is not evaluated to a value to be bound, but changes the machine state through the indices (e.g. $\kw{r}\in\kw{greg}$) -- to the global maps. 
\subsection{Instructions}
\label{sec:instrsemantics}
The syntax of instructions is given in \fref{fig:syntax}. Their
small-step operational semantics, a~reduction relation of the form
$\hsfork{\instr}{\store}{\instr'}{\store'}{\instrs}$,
is defined in \fref{fig:semantics}.
%
The last argument~$\instrs$ is a list of newly-spawned threads;
when it is empty, we omit it
and write just $\hs{\instr}{\store}{\instr'}{\store'}$.
\subsection{A Set of Selected Instructions Changing the Machine State}
\label{sec:selectedinstrsemantics}
In addition to the reduction rules in Figure \fref{fig:reductionsemantics} for the light-weight integration of our machine model to our simple language, and, finally, to our reasoning principles, we think that it is worth building the intuition on the way some of the instructions -- specifically the ones used in the examples contained in this paper -- change the machine state when reduced with \RULE{StepContext}. \RULE{StepMovRR} in \fref{fig:instrsemantics} shows how machine state change when we copy the source register's value ($\rvsrc$) referenced by the register name ($\rgsrc$) to the destination register ($\rgdst$). We follow the convention of showing only the pieces of the machine state changed after the instruction execution -- only the register value of the destiantion register ($ \readlval\rgdst\storeregprime\rvsrc$) has changed in the state ($\storeprime$) after the execution of the instruction $\textsf{mvrr}$. All the other rules in Figure \fref{fig:instrsemantics}, presents the change in the state when a value is loaded/stored to/from register/memory, and are differentiated with respect to the addressing mode ($amode$) which determines whether the accessed memory address is computed based on some offset (\RULE{StepMovRMOff}, \RULE{StepMovMROff}) or not (\RULE{StepMovRMBase}) and \RULE{StepMovMRBase}). The most common essential aspect of loading the source register's value ($\rvsrc$) to the memory, \RULE{StepMovRMBase} and \RULE{StepMovRMOff}, is the hierarchical translation of memory location, $\readlval{\plusaddr\maddr\offs}{\storememstar\crval}\locsf$, which is hidden in the rules, expanded as the following:
\begin{itemize}
\item
\item
\item
\item
\end{itemize}
where the first three levels of translation allows address-translation .
\input{fig-semantics}

>>>>

\subsection{State}
\label{sec:state}
\todo[inline]{I don't think we need or want to describe the operational semantics in detail in the submission. It adds a lot of formalization that doen'st necessarily contribute to reader understanding. We should describe the runtime \emph{state} $\sigma$ of memory and registers, since that is directly used in interpreting assertions. But the reduction rules will mostly add mental overhead here.}
We represent the machine state mainly as a finite map of registers to register values and a map of masked physical memory addresses to 64-bit physical memory values. Together with default def\textsf{CPU} instantiation, we end up having our state with following pieces:
\begin{itemize}
\item A constant cpu initialized with default value: $\sigma.\mathcal{C}$
\item A register map: $\sigma.\mathcal{R}: \kw{greg} \rightarrow_{\textrm{fin}} \kw{regval} $
\item A memory map: $\sigma.\mathcal{M}: \Locft \rightharpoonup_{\textrm{fin}} (\Loctw \rightharpoonup_{\textrm{fin}} \Locsf )$
\end{itemize}
\subsection{Memory Addresses \& Naming Registers}
\label{sec:}
As one might have already anticipated from the syntax we introduce, we do not bind any value of an evaluated expression. All the indices and accessed values are treated as globally referenced. In align with this design choice, our expression is a stream of instructions, which is not evaluated to a value to be bound, but changes the machine state through the indices (e.g. $\kw{r}\in\kw{greg}$) -- to the global maps. 
\subsection{Instructions}
\label{sec:instrsemantics}
The syntax of instructions is given in \fref{fig:syntax}. Their
small-step operational semantics, a~reduction relation of the form
$\hsfork{\instr}{\store}{\instr'}{\store'}{\instrs}$,
is defined in \fref{fig:semantics}.
%
The last argument~$\instrs$ is a list of newly-spawned threads;
when it is empty, we omit it
and write just $\hs{\instr}{\store}{\instr'}{\store'}$.
\subsection{Machine State under Address Translation}
\label{sec:selectedinstrsemantics}
Althought we give the complete set of operational semantics rule in \sref{appendix:movops}, It is worth building the intuition on the the way some of these rules bahave in the context of address translation. In fact $\readlval\maddr{\storememstar\crval}\locsf$ is unfoled into multiple physical memory lookups for the final page address retrieval -- i.e. address translation traversal as shown in Figure \ref{fig:pagetables}:
\begin{itemize}
\item top-level-address translation: with the given root address (64-bit $\crval$) of the address space, performs the address translation, handling the first level (to get the PML4 entry) itself. The next level table address is computed with the fetched PML4 offset value which exhibits itself as 9-bit offset in $\kw{maddr}$. \mytodo{iso: put move}
\item translating from PML4 entry: performs the second level of address translation, to retrieve starting at the PML4 table entry, and interprets the PML4 entry that references a Page-Directory-Pointer Table (PDPT). We obtain the PDPT offset which exhibits itself as 9 bit offset in $\kw{maddr}$ to obtain the address of the next level page directory table (PDT) \mytodo{iso: put move}
\item translating from PD entry: performs the third level of address translation, to retrieve starting at the PDP table entry, and interprets a PDPTE that references a PD table. Likewise, we obtain the PD offset which exhibits itself as 9 bit offset in $\kw{maddr}$ to obtain the address of the next level page directory table (PT) \mytodo{iso:put move }
\item translate from PT entry: performs final level of address translation, starting from the PT entry, with a given 12 bit page offset, we can compute the physical address referencing $\locsf$ \mytodo{ismo: put mov}
\end{itemize}
